Dynamical properties of the Benguela Current as seen through a cold-spell algorithm.

### 2019-03-03 Write_up/Writing/dissertation.Rmd
updated

## Eastern boundary upwelling systems:

Globally, there are four major eastern boundary upwelling systems (EBUS). These are the Benguela Current off south-western Africa, the Canary Current off north-western Africa including its northern extension off the Iberian Peninsula of south-western Europe, the Peru-Humboldt Current off western South America, and the California Current off the western continental USA and north-western Mexico [@bakun ; @pauly]. Each of these systems are characterized as vast regions of coastal ocean occurring along the western shores of continents bordering the Pacific and Atlantic Oceans. EBUS cover only 1% of the world’s ocean surface, but despite these are among the most productive regions of the world [@pauly]. This is because these systems are subject to coastal upwelling, which is defined as the process whereby cold, nutrient rich, high CO2 concentrated, low pH, and low oxygen waters are pushed to the surface as a result of alongshore winds interacting with the earth’s rotation [@huyer]. As a result of upwelling, diverse marine fauna and flora are able to thrive in these areas because of the abundance of nutrients and favourable temperatures.

Upwelling is primarily caused by alongshore, equator ward winds. These winds are caused by cross-shore atmospheric pressure gradients, and these gradients occur predominantly during heating periods. Bakun then proposed that if there is an increase in greenhouse gases, it will cause more heating during the day and more cooling during the night which cause the temperature gradient to increase and would result in a strong pressure gradient, and as we know the pressure gradient will affect the wind and therefore will affect upwelling. Several people have tried to model this using different datasets of SST and wind pattern data at various resolution and initially found that increases in gases, such as CO^2 causes a reduction in trade winds, and therefore a reduction in upwelling. However, these models were based on coarse data and when higher resolutions are used, such as for the Benguela EBUS, the models have shown that upwelling will intensify in accordance to the upwelling intensification hypotheses [@hsieh; @mote; @lima].

### Ecology and primary production within upwelling zones:

The duration and intensity of coastal upwelling are known to have a critical role in the phenology of key marine ecosystem processes. Similarly, changes in the characteristics of upwelling are shown to have cause large disturbances to ecosystems at several trophic levels [@wang]. Cold, nutrient rich, low pH, and low oxygen upwelled waters fuels primary production as a consequence of the development of phytoplankton blooms [@ryther; @Cushing; @mcGillicuddy; @oschlies]. In these upwelling systems, the high productivity supports vast and diverse marine populations. Mesoscale and sub-mesoscale oceanic movements are key in managing the exchange and transportation of organic resources and marine organisms from the coastal zone to the open ocean. Most of the world’s major fishing industries are dependent on upwelling, because this process brings nutrient rich waters into the euphotic zone, where it forms the foundation of a localized food web that ecologically differs from surrounding, non-upwelled regions [@bosman; @barshis]. Whilst nutrients play a primary role in the survival of these systems, the cool temperatures of these waters are critical as well. There is an increasingly large body of research that has provided evidence showing that sea surface temperatures (SST) have and are continuing to increase. However, some regions still show a lower or non-significant increase, and in some cases cooling trends are evident (REFERENCES). Upwelling occurring along the Cape south coast is proven to have less of an economic importance in contrast to the west coast. Inshore regions and embayments that are protected from wind action may be subjected to intense episodes of primary production [@graham]. Vander Wouder et al. (2006) found that these sites represent the highest levels of chlorophyll concentration. Biotic systems within the Benguela upwelling system may be negatively influenced by the changes in oxygen levels given their adaption to natural variable conditions [@capone].

For several decades, regime shifts have become an important factor of marine research. Regime shifts are often defined as a shift from one stable ecosystem state to another. This process affects several species within a short period of time [@cury; @jarre; @blamey]. Regime shifts are thought to be driven by abiotic processes such as climate change [@polovina; @blamey]. The southern Benguela upwelling system supports a wide variety of commercially-exploited fish species and exhibits substantial variability in both oceanographic and biological components. Over the past 3 three decades, a number of species occurring within this region have undergone shifts in their distributions and Roy et al. (2007) linked these shifts to cooling waters along the inner shelf of the Agulhas bank on the south coast, east of Cape Agulhas [@blamey]. This coast is associated with high levels of harmful toxic algal blooms, which negatively impact commercial and recreational interests along the coastal region. These harmful blooms are often attributed to flagellated species with the red tide representing dinoflagellate blooms (Pitcher et al. 1993A, @pitcher1998]. Formations of red tides are often related to the prevailing winds of the southern Benguela which govern most hydrodynamic processes on the continental shelf. Red tide events are greatest during the latter part of the upwelling season during transitions in the synoptic weather patterns which result in diminished upwelling activity and increased thermal stratification (Pitcher et al. 1993a).

### Atmosphere:

Throughout region surrounding the Benguela EBUS, the atmosphere is dominated by anticyclonic high pressure cells with quasi-stationary positions. The South Atlantic Ocean High is situated along the west, drawing cool, dry air onto the west of the subcontinent [@van]. Solar heating during summer may result in the development of low pressure cells, known as heat lows. These cells are absent during winter time [@tyson]. When the anticyclones shift north, during winter months, the cold westerlies have a substantial impact on the weather of the southern tip of the South African subcontinent [@van]. The atmospheric temperatures along the coastline are largely influenced by the Benguela Current [@van]. However, upwelling systems may be forced by various atmospheric conditions [@Lutjeharms; @roberts; @hutchings].

### Sea surface temperature and data collection:

Water temperature affects the survival, abundance, distribution and growth of marine species (REFERENCE). As such, a clear understanding of the relationship between marine ecosystems and their thermal tolerance is necessary for conservation and management planning. Future predictions in biogeographical patterns and ecosystem functioning may also be estimated based on various climatic behaviours. Regardless of the objectives, whether it be studies based on climate change, extreme events or ecosystem functioning, it is important that the research is conducted through accurate, reliable datasets. Unfortunately, not all data are created equally and under the same conditions and as such not all datasets are usable for a particular task.

Remotely-sensed satellite SST products are produced and maintained by a number of organisations worldwide. In order to obtain these SST data, advanced technology is launched into space, in order to obtain infra-red light being emitted by the ocean. This irradiance is then taken as a proxy for SST.  Due to the distance at which these SST are being collected, certain coarseness within the datasets are unavoidable. Satellite SST products undergo a range of quality control processes that are well researched and tested [@reynolds], making their use more reliable. However, Smit et al. (2013) have shown that SST data are not particularly reliable for coastal regions and should be used cautiously as they can be prone to both warm and cold bias. Nevertheless, satellite data products are extremely useful when investigating the ocean at broad scale and can provide important SST data.

@@

Having coastal, in situ seawater temperature data is a rare privilege to researchers. Most biological and ecological research requires coastal temperature data. The South African Coastal Temperature Network (SACTN). The availability of the in situ collected coastal temperature data products for several decades and provides a reliable source of accurate coastal seawater temperature data [@smit]. SACTN has collected coastal seawater temperature along the South African coastline from as early as 1972, with contributions from a total of seven different organisations and governmental departments. At present, this dataset contained temperature of 129 sites along the three distinct coasts of the South African coastline. Temperature recordings are still ongoing at many of these sites, but unfortunately, the durations and extents of these are uneve per site. Data collected for this region started on 13 September 1972 and concluded on 26 January 2017, with recordings still continuing daily. During the 1970s, a total of 11 time series began recording. A further 53 entries were added during the 1980s, 34 entries were added during the 1990s, and 18 entries were added during the 2000s.

The uneven and inconsistent nature of the SATCN dataset can be attributed to several factors. For instance, these may be as a result of variances in the equipment used by the various organisations and governmental bodies who recorded the data. One problematic outcome of this sampling methodology is that due to specific work of the South African Weather Service (SAWS) and Kwazulu Natal Sharks Board (KZNSB), two thirds of the SACTN data that were sampled were obtained with hand-held thermometers and records at a data precision of 0.500 ◦ C. However, the current global standard is to make use of underwater temperature recorders (UTRs), which have a data precision of 0.001◦ C [@jarraud]. Because of these inconsistencies, this dataset required verification from an independent source before its use within research. Smit et al. (2013) thoroughly compared the temperature data within the SACTN dataset to those of several remotely-sensed SST products and found bias results. Using these in situ collected temperature data along with remotely sensed SST data requires a clear understanding in the characteristics of the contributing data sources in order to understand how accurate, reliable and useful these long term measurements are for use in climate change studies.

It must also be acknowledged that most of the time series within the SACTN dataset do not have meta-data records that meet the international climate change research requirements [@aguilar]. Specifically uncertainty lies in the drift of the instrument, who was sampling and where the exact sampling locations were collected throughout the collection period. We are however certain that all of the thermometer time series were collected by thermometers and that the UTR time series were collected by UTRs. Given these issues associated with the mete-data, recent work has been started to develop a nation wide standard.

@@

Satellite SST products undergo a range of quality control and validation processes (e.g. Reynolds and Smith, 1994; Brown et al., 1999; Martin et al., 2012), making the data reliable and more useful. As previously mentioned, satellites may show a warm or cold bias along the coastal region but they are still useful for broader scale investigations (Smit et al., 2013, Smale and Wernberg, 2009; Castillo and Lima, 2010; Dufois and Rouault). Remotely-sensed SST products are obtained at different resolutions, and as a result do not show the same level of bias against in situ collected seawater temperature. Research done by Smit et al (2013) showed that MODIS represented the least amount of bias.

This study made use of various satellite-derived Sea Surface Temperatures (SST) datasets. One of these datasets are the National Oceanic and Atmospheric Administration (NOAA) ¼° daily Optimum Interpolation Sea Surface Temperature (OISST). This dataset is an analysis of seawater temperature by different platforms (satellites, ships and buoys) on a global grid. The gridded daily SST product assimilates both remotely sensed and *in situ* sources of data to create a level-4 gap free product (Banzon et al., 2016). A spatially complete sea suface temperature map is produced by interpolation to fill in gaps within the dataset. The method used to fill these gaps includes the adjustment of satellite and ship observations which compensate for the bias differences (NOAA.GOV, https://www.ncdc.noaa.gov/oisst). There are two types of daily OISST, named after the relevant satellite sensors. These are the Advanced Very High Resolution Radiometer (AVHRR) and Advanced Microwave Scanning Radiometer on the Earth Observing System (AMSR-E). Currently the AVHRR has the longest record starting from 1981 to the present. Infrared instruments, like AVHRR, can make observations at relatively high resolutions. Microwave instruments, like AMSR-E, can measure SSTs in most weather conditions but not adjacent to land. The AVHRR-only Optimally-Interpolated Sea Surface. Temperature (OISST) was used to determine SST within the study region. The AVHRR datasets have been provided global SSTs for more than four decades [@reynolds; @Pearce; @corlett].

The Multi-scale Ultra-high Resolution (MUR) Sea Surface Temperature Analysis, the second dataset, is produced using satellite instruments with datasets spanning 1 June 2002 to present times (PO.DAAC, https://podaac.jpl.nasa.gov/). MUR provides SST data at a spatial resolution of 0.01° in longitude-latitude coordinates, roughly at 1 km intervals. This dataset is currently among the highest resolution SST datasets available. The goal in creating the MUR SST is to develop a coherent and consistent daily map of SST at the highest spatial resolution possible. The MUR sea surface temperature dataset combines data from the Advanced Very High Resolution Radiometer (AVHRR), Moderate Imaging Spectroradiometer (MODIS) Terra and Aqua, and Advanced Microwave Spectroradiometer-EOS (AMSR-E) instruments in an optimal way to produce 1km global maps representing the sea surface temperature.

The Moderate-resolution Imaging Spectroradiometer (MODIS) is a scientific instrument (radiometer) on board the NASA Terra and Aqua satellite platforms with a spatial resolution in 0.041° longitude-latitude coordinates. MODIS was launched in 1991 and 2002 to study global dynamics of the earths atmosphere, ice and ocean. MODIS captures data in 36 spectral bands ranging in wavelength from 0.4 um to 14.4 um and at varying spatial resolutions (2 bands at 250 m, 5 bands at 500 m and 29 bands at 1 km), providing complete global coverage of the Earth every one to two days [@hosoda; @gentemann, @dong]. MODIS Terra Global level 3 mapped thermal SST products consists of sea surface temperature data derived from the 11 and 12um thermal IR infrared (IR) Bands (NASA.GOV, http://oceancolor.gsfc.nasa.gov).

The CMC dataset constitutes the fourth dataset and is a version 3.0 Group for High Resolution Sea Surface Temperature (GHRSST) Level 4 dataset with a 10 km resolution constructed by the Canadian Meteorological Center (CMC). The CMC dataset combines infrared satellite SST at numerous points in the time series from the AVHRR, the European Meteorological Operational-A (METOP-A) and Operational-B (METOP-B) platforms, and microwave SST data from the Advanced Microwave Scanning Radiometer 2 in conjunction with in situ observations of SST from ships and buoys from the ICOADS program.

The fifth dataset, the Global 1 km SST (G1SST) analysis, representing a 0.01° resolution, uses satellite data from sensors that include the Advanced Very High Resolution Radiometer (AVHRR), the Advanced Along Track Scanning Radiometer (AATSR), the Spinning Enhanced Visible and Infrared Imager (SEVIRI), the Advanced Microwave Scanning Radiometer-EOS (AMSRE), the Tropical Rainfall Measuring Mission Microwave Imager (TMI), the Moderate Resolution Imaging Spectroradiometer (MODIS), the Geostationary Operational Environmental Satellite (GOES) Imager, the Multi-Functional Transport Satellite 1R (MTSAT-1R) radiometer, and in situ data from drifting and moored buoys [@dash}.  Thid dataset produces a product that combines satellite data with the model output (PO.DAAC, https://podaac.jpl.nasa.gov/).

### Climate change:

Climate change as a result of anthropogenic activities is  an established global concern of both scientific and political importance. As the human population grows, the rate of global warming increases, resulting in profound effects on marine and other ecosystems. Climbing temperatures create a host of additional changes to marine systems, such as rising sea levels, increased ocean stratification, decreased sea-ice extent, and altered patterns of ocean circulation and precipitation [@doney]. Furthermore, direct effects on ocean temperature changes and chemistry may alter the functioning and behaviour or marine organisms. In turn, these changes lead to a change in species integration and trophic pathways.

Significant research studies accept the climate change and variability have great impacts on marine ecosystems [@stenseth; @harley; @hoegh]. The Intergovernmental Panel on Climate Change (IPCC) predicts future projections with a mean increase in air temperature of 1.1°C by the 2020’s and 1.8°C by the 2040’s as a result of greenhouse gas concentrations [@solomon]. A large amount of this heat will be absorbed by the world’s ocean. In current years, as a result of changes in air temperature, the heat content of the ocean has increased largely. Given ocean warming in association with their physical changes such as sea level rise and declines in sea ice, marine life is expected to intensify in future decades. Global warming represents distinct warming and cooling periods throughout the 20th and 21st century (IPCC, 2007). Global warming however is not evenly distributed across the global ocean, and thus several regions are warming faster or slower than then García-Reyes et al., 2015 the global average [@levitus; @santos]. In particular, the Atlantic Ocean contributes most to the increase of the heat content [@nerem; @levitus].

As a result of climate change, research studies provide significant evidence relating to the increase in the warming of the ocean surface. This warming slowly propagates down through the water column and in turn increases the thermal stratification [@gruber]. Strong stratification limits the depth at which water is upwelled and hence controls the amount of nutrients brought up [@chhak; @jacox2011; @jocox2016]. The interplay between upwelling winds and stratification determines the amount of nutrients present within the system and is therfore strongly linked to biological productivity. Globally, as a result of climate change, stratification of the water-column between 0 and 200 m has increased by an estimated 4% between 1971 to 2010 [@stocker]. Increased ocean stratification as a result of surface warming may reduce the ventilation of deep-water masses. As the deep water is being upwelled to the surface, decreases in oxygen and pH levels are predicted along with an increase in nutrient levels [@rykaczewski].  Generally, regional ocean climate oscillations like the Benguela Niño for the Benguela Current System impose considerable variability on temperatures and complicate attempts to assess long-term trends [@demarcq]. It is predicted that these climate oscillations may increase the variance with climate change [@timmermann; @kuzmina; @sydeman] which will further result in unusual trends.

@@

Climate change has severe ecological consequences on both plants and animals (Bradley et al. 1999; Walther et al. 2002). For example, in the past two decades, sea surface temperature has risen in many regions. Seasonal rise in temperature can act as an important cue in species survival, reproduction and abundance (Cushing 1982; Cushing 1990). The differing responses of species as a result of climate change are important to understand as these changes affect, among many things, the trophic dynamics and fisheries yield of several regional ecosystems (Murawski 1993). Understanding the relationship between global warming and climate change as a result of anthropogenic activity is a vast and challenging process. In a bid to further this understanding, Rockström et al. (2009) introduced the concept of planetary boundaries. According to this concept, within each of these boundaries exists a level of safety that should not be exceeded. For example, the planetary boundary for biodiversity loss, has been surpassed [@steffen]. Similarly, the safety boundary for warming of the globe by greenhouse gas emission has also been surpassed. The idea of planetary boundaries is essential in that it helps our understanding of the threats we cause via the continued destruction of the natural environment. Consequently, studies of anthropogenic influences on the ecosystem are generally divided into land or sea. Terrestrial ecosystems are more familiar to the general populace; whereas marine ecosystems are associated with high levels of primary production [@Chavez] and are highly at risk as a result of climate change [@osterblom]. Climate change has been proven to decrease primary production thus affecting the trophic levels and leading to trophic turbulence and destruction of marine fauna [@lewandowska; @dell, @mcCauley].

Climate change similarly affects coastal species such as oysters, corals and, sea grass. Lord (2017) also found evidence that suggests that a change in climate as a result of rising sea temperatures may increase the spread of invasive species and increase the chances of an outbreak of diseases [@altizer]. Research by Mead et al (2013) found that coastal ecosystems are at high risk as a result of climate change, later, Whitfield et al (2016) confirmed that coastal bio-diversity have been observed to be affected. Given the importance of primary productivity, it was determined that the Benguela upwelling system within the South African region should become the focus of this thesis. Here the increase in global temperatures are one of the main focus points along with the chlorophyll concentration. By making use of these two measurements, I am able to analyses how chlorophyll concentration vary with a change in temperature.

## Benguela :

>>>>

## Eastern boundary upwelling systems

Globally, there are four major eastern boundary upwelling systems (EBUS). Upwelling systems include the Benguela Current off south-western Africa, the Canary Current off north-western Africa together with its northern extension off the Iberian Peninsula of south-western Europe, the Peru-Humboldt Current off western South America, and the California Current off the western continental USA and north-western Mexico [@bakun ; @pauly]. Each of these systems are characterized as vast regions of coastal ocean occurring along the western shores of continents bordering the Pacific and Atlantic Oceans. EBUS cover only 1% of the world’s ocean surface, but despite this these are among the most productive regions of the world [@pauly]. This is due to the  systems exposure to coastal upwelling, which is defined as the process whereby cold, nutrient rich, high concentrated CO2, low pH, and low oxygenated waters are pushed to the surface as a result of alongshore winds interacting with the earth’s rotation [@huyer]. As a result of upwelling, diverse marine fauna and flora are able to thrive in these areas due to the abundance of nutrients and favourable temperatures.

Upwelling is primarily caused by alongshore, equator ward winds. These winds are caused by cross-shore atmospheric pressure gradients, and these gradients occur predominantly during heating periods. Bakun proposed that if there was an increase in greenhouse gases, it would cause more heating during the day and more cooling during the night causing the temperature gradient to increase resulting in a strong pressure gradient, Ultimately effecting wind patterns and therefore upwelling. Several people have tried to model this using different datasets of SST and wind pattern data at various resolutions and initially found that increases in gases, such as CO^2 causes a reduction in trade winds, and therefore a reduction in upwelling. These models however, were based on coarse data and when higher resolutions were used, such as for the Benguela EBUS, the models have shown that upwelling will intensify in accordance to the upwelling intensification hypotheses [@hsieh; @mote; @lima].

### Ecology and primary production within upwelling zones

The duration and intensity of coastal upwelling are known to play a critical role in the phenology of key marine ecosystem processes. Similarly, changes in the characteristics of upwelling are shown to cause large disturbances to ecosystems at several trophic levels [@wang]. Cold, nutrient rich, low pH, and low oxygen upwelled waters fuels primary production as a consequence of the development of phytoplankton blooms [@ryther; @Cushing; @mcGillicuddy; @oschlies]. In these upwelling systems, the high productivity supports vast and diverse marine populations. Mesoscale and sub-mesoscale oceanic movements are key in managing the exchange and transportation of organic resources and marine organisms from the coastal zone to the open ocean. Most of the world’s major fishing industries are dependent on upwelling, as this process brings nutrient rich waters into the euphotic zone, where it forms the foundation of a localized food web that ecologically differs from surrounding, non-upwelled regions [@bosman; @barshis]. Whilst nutrients play a primary role in the survival of these systems, the cool temperatures of these waters are critical as well. There is an increasingly large body of research that has provided evidence showing that sea surface temperatures (SST) have and are continuing to increase, however some regions still show a lower or non-significant increase, and in some cases cooling trends are evident (REFERENCES). Upwelling occurring along the Cape south coast is proven to have less of an economic importance in contrast to the west coast. Inshore regions and embayments that are protected from wind action are likely to be subjected to intense episodes of primary production [@graham]. Vander Wouder et al. (2006) found that these sites represent the highest levels of chlorophyll concentration. Biotic systems within the Benguela upwelling system may be negatively influenced by the changes in oxygen levels given their adaption to natural variable conditions [@capone].

For several decades, regime shifts have become an important factor of marine research. Regime shifts are often defined as a shift from one stable ecosystem state to another. This process affects several species within a short period of time [@cury; @jarre; @blamey]. Regime shifts are thought to be driven by abiotic processes such as climate change [@polovina; @blamey]. The southern Benguela upwelling system supports a wide variety of commercially-exploited fish species and exhibits substantial variability in both oceanographic and biological components. Over the past  three decades, a number of species occurring within this region have undergone shifts in their distributions and Roy et al. (2007) associated these shifts to cooling waters along the inner shelf of the Agulhas bank on the south coast, east of Cape Agulhas [@blamey]. This coast is associated with high levels of harmful toxic algal blooms, which negatively impact commercial and recreational interests along the coastal region. These harmful blooms are often attributed to flagellated species with the red tide representing dinoflagellate blooms (Pitcher et al. 1993A, @pitcher1998]. Formations of red tides are often related to the prevailing winds of the southern Benguela which govern most hydrodynamic processes on the continental shelf. Red tide events are greatest during the latter part of the upwelling season during transitions in the synoptic weather patterns which result in diminished upwelling activity and increased thermal stratification (Pitcher et al. 1993a).

### Atmosphere

Throughout regions surrounding the Benguela EBUS, the atmosphere is dominated by anticyclonic high-pressure cells with quasi-stationary positions. The South Atlantic Ocean High is situated along the west, drawing cool, dry air onto the west of the subcontinent [@van]. Solar heating during summer may result in the development of low-pressure cells, known as heat lows, which are absent during the winter [@tyson]. As the anticyclones shift north during winter months, the cold westerlies have a substantial impact on the weather of the southern tip of the South African subcontinent [@van]. The atmospheric temperatures along the coastline are largely influenced by the Benguela Current [@van], however upwelling systems may be forced by various atmospheric conditions [@Lutjeharms; @roberts; @hutchings].

### Sea surface temperature and data collection

Water temperature affects the survival, abundance, distribution and growth of marine species (REFERENCE). As such, a clear understanding of the relationship between marine ecosystems and their thermal tolerance is necessary for conservation and management planning. Future predictions in biogeographical patterns and ecosystem functioning may also be estimated based on various climatic behaviours. Regardless of the objectives, whether it be studies based on climate change, extreme events or ecosystem functioning, it is important that the research is conducted through accurate, reliable datasets. Unfortunately, not all data are created equally and under the same conditions and as such, not all datasets are appropriate for a particular task.

@@

Having coastal, in situ seawater temperature data is a rare privilege to researchers. Most biological and ecological research requires coastal temperature data. The South African Coastal Temperature Network (SACTN). The availability of the in situ collected coastal temperature data products for several decades provides a reliable source of accurate coastal seawater temperature data [@smit]. SACTN has collected coastal seawater temperature along the South African coastline from as early as 1972, with contributions from a total of seven different organisations and governmental departments. At present, this dataset contains temperatures of 129 sites along the three distinct coasts of the South African coastline. Temperature recordings are still ongoing at many of these sites, but unfortunately, the durations and extents of these are uneven per site. Data collected for this region started on 13 September 1972 and concluded on 26 January 2017, with recordings still continuing daily. During the 1970s, a total of 11 time series began recording. A further 53 entries were added during the 1980s, 34 entries were added during the 1990s, and 18 entries were added during the 2000s.

The uneven and inconsistent nature of the SATCN dataset can be attributed to several factors. For instance, as a result of variances in the equipment used by the various organisations and governmental bodies who recorded the data. One problematic outcome of this sampling methodology was that, due to specific work of the South African Weather Service (SAWS) and Kwa-Zulu Natal Sharks Board (KZNSB), two thirds of the SACTN data that were sampled were obtained with hand-held thermometers and recorded at a data precision of 0.500 ◦ C, however, the current global standard is to make use of underwater temperature recorders (UTRs), which have a data precision of 0.001◦ C [@jarraud]. Due of these inconsistencies, this dataset required verification from an independent source before its use in research. Smit et al. (2013) thoroughly compared the temperature data within the SACTN dataset to those of several remotely-sensed SST products and found bias results. Using these in situ collected temperature data along with remotely sensed SST data requires a clear understanding of the characteristics of the contributing data sources in order to understand how accurate, reliable and useful these long-term measurements are for use in climate change studies.

It should also be acknowledged that most of the time series within the SACTN dataset do not have meta-data records that meet the international climate change research requirements [@aguilar]. Specifically, uncertainty lies within the drift of the instrument, who was sampling and where the exact sampling locations were collected throughout the collection period. There is certainty that all of the thermometer time series were collected by thermometers and that the UTR time series were collected by UTRs in this study.  Given these issues associated with the mete-data, recent work has been started to develop a nation wide standard.

@@

Satellite SST products undergo a range of quality control and validation processes (e.g. Reynolds and Smith, 1994; Brown et al., 1999; Martin et al., 2012), making the data reliable and more useful. As previously mentioned, satellites may show a warm or cold bias along the coastal region but are still useful for broader scale investigations (Smit et al., 2013, Smale and Wernberg, 2009; Castillo and Lima, 2010; Dufois and Rouault). Remotely-sensed SST products are obtained at different resolutions, and as a result do not represent the same level of bias against in situ collected seawater temperature. Research done by Smit et al (2013) showed that MODIS represented the least amount of bias.

This study made use of various satellite-derived Sea Surface Temperature (SST) datasets. One of datasets used was the National Oceanic and Atmospheric Administration (NOAA) ¼° daily Optimum Interpolation Sea Surface Temperature (OISST). This dataset is an analysis of seawater temperature collected by different platforms (satellites, ships and buoys) on a global grid. The gridded daily SST product assimilates both remotely sensed and *in situ* sources of data to create a level-4 gap free product (Banzon et al., 2016). A spatially complete sea surface temperature map is produced by interpolation to fill in gaps within the dataset. The method used to fill these gaps include the adjustment of satellite and ship observations which compensate for bias differences (NOAA.GOV, https://www.ncdc.noaa.gov/oisst). There are two types of daily OISST, named after the relevant satellite sensors, these are the Advanced Very High Resolution Radiometer (AVHRR) and Advanced Microwave Scanning Radiometer on the Earth Observing System (AMSR-E). Currently the AVHRR has the longest record starting from 1981 to present day. Infrared instruments, like AVHRR, can make observations at relatively high resolutions. Microwave instruments, such as AMSR-E, can measure SSTs in most weather conditions but not adjacent to land. The AVHRR-only Optimally-Interpolated Sea Surface Temperature (OISST) was used to determine SST within the study region. The AVHRR datasets have provided global SSTs for more than four decades [@reynolds; @Pearce; @corlett].

The Multi-scale Ultra-high Resolution (MUR) Sea Surface Temperature Analysis, used as the second dataset, was produced using satellite instruments with datasets spanning from1 June 2002 to present times (PO.DAAC, https://podaac.jpl.nasa.gov/). MUR provides SST data at a spatial resolution of 0.01° in longitude-latitude coordinates, roughly at 1 km intervals. This dataset is currently among the highest resolution SST dataset available. The goal in creating the MUR SST is to develop a coherent and consistent daily map of SST at the highest spatial resolution possible. The MUR sea surface temperature dataset combines data from the Advanced Very High Resolution Radiometer (AVHRR), Moderate Imaging Spectroradiometer (MODIS) Terra and Aqua, and Advanced Microwave Spectroradiometer-EOS (AMSR-E) instruments in an optimal way in order to produce 1km global maps representing sea surface temperature.

The Moderate-resolution Imaging Spectroradiometer (MODIS) is a scientific instrument (radiometer) on board the NASA Terra and Aqua satellite platforms with a spatial resolution in 0.041° longitude-latitude coordinates. MODIS was launched in 1991 and 2002 to study global dynamics of the earth’s atmosphere, ice and ocean. MODIS captures data in 36 spectral bands ranging in wavelength from 0.4 um to 14.4 um and at varying spatial resolutions (2 bands at 250 m, 5 bands at 500 m and 29 bands at 1 km), providing complete global coverage of the Earth every one to two days [@hosoda; @gentemann, @dong]. MODIS Terra Global level 3 mapped thermal SST products consists of sea surface temperature data derived from the 11 and 12um thermal IR infrared (IR) Bands (NASA.GOV, http://oceancolor.gsfc.nasa.gov).

The CMC dataset comprises the fourth dataset and is a version 3.0 Group for High Resolution Sea Surface Temperature (GHRSST) Level 4 dataset with a 10 km resolution constructed by the Canadian Meteorological Center (CMC). The CMC dataset combines infrared satellite SST at numerous points in the time series from the AVHRR, the European Meteorological Operational-A (METOP-A) and Operational-B (METOP-B) platforms, and microwave SST data from the Advanced Microwave Scanning Radiometer 2 in conjunction with in situ observations of SST from ships and buoys from the ICOADS program.

The fifth dataset, the Global 1 km SST (G1SST) analysis, representing a 0.01° resolution, uses satellite data from sensors that include the Advanced Very High Resolution Radiometer (AVHRR), the Advanced Along Track Scanning Radiometer (AATSR), the Spinning Enhanced Visible and Infrared Imager (SEVIRI), the Advanced Microwave Scanning Radiometer-EOS (AMSRE), the Tropical Rainfall Measuring Mission Microwave Imager (TMI), the Moderate Resolution Imaging Spectroradiometer (MODIS), the Geostationary Operational Environmental Satellite (GOES) Imager, the Multi-Functional Transport Satellite 1R (MTSAT-1R) radiometer, and in situ data from drifting and moored buoys [@dash]. This dataset produces a product that combines satellite data with the model output (PO.DAAC, https://podaac.jpl.nasa.gov/).

### Climate change

Climate change as a result of anthropogenic activities is a global concern of both scientific and political importance. As the human population grows, the rate of global warming increases, resulting in profound effects on marine and other ecosystems. Climbing temperatures create a host of additional changes to marine systems, such as rising sea levels, increased ocean stratification, decreased sea-ice extent, and altered patterns of ocean circulation and precipitation [@doney]. Direct effects on ocean temperature changes and chemistry may alter the functioning and behaviour of marine organisms and in turn, these changes lead to a change in species integration and trophic pathways.

Significant research studies accept that climate change and variability have great impacts on marine ecosystems [@stenseth; @harley; @hoegh]. The Intergovernmental Panel on Climate Change (IPCC) predicts future projections with a mean increase in air temperature of 1.1°C by the 2020’s and 1.8°C by the 2040’s as a result of the increase of greenhouse gas concentrations [@solomon]. A large amount of this heat will be absorbed by the world’s ocean. In current years, as a result of changes in air temperature, the heat content of the ocean has increased significantly. Given ocean warming in association with their physical changes such as sea level rise and declines in sea ice, marine life is expected to intensify in future decades. Global warming represents distinct warming and cooling periods throughout the 20th and 21st century (IPCC, 2007). Global warming however is not evenly distributed across the global ocean, and thus several regions are warming faster or slower than García-Reyes et al., 2015 the global average [@levitus; @santos]. In particular, the Atlantic Ocean contributes most to the increase of heat content [@nerem; @levitus].

As a result of climate change, research studies provide significant evidence relating to the increase in the warming of the ocean surface. This warming slowly propagates down through the water column and in turn increases thermal stratification [@gruber]. Strong stratification limits the depth at which water is upwelled and hence controls the amount of nutrients brought up [@chhak; @jacox2011; @jocox2016]. The interplay between upwelling winds and stratification determines the amount of nutrients present within the system and is therefore strongly linked to biological productivity. Globally, as a result of climate change, stratification of the water-column between 0 and 200 m has increased by an estimated 4% between 1971 to 2010 [@stocker]. Increased ocean stratification as a result of surface warming may reduce the ventilation of deep-water masses. As the deep water is being upwelled to the surface, decreases in oxygen and pH levels are predicted with an increase in nutrient levels [@rykaczewski].  Generally, regional ocean climate oscillations like the Benguela Niño for the Benguela Current System impose considerable variability on temperatures and complicate attempts to assess long-term trends [@demarcq]. It is predicted that these climate oscillations may increase the variance with climate change [@timmermann; @kuzmina; @sydeman] which may result in unusual trends.

@@

Climate change has severe ecological consequences on both plants and animals (Bradley et al. 1999; Walther et al. 2002). In the past two decades, sea surface temperature has risen in many regions and this Seasonal rise in temperature can act as an important cue in species survival, reproduction and abundance (Cushing 1982; Cushing 1990). The different responses of species as a result of climate change are important to understand as these changes affect, the trophic dynamics and fisheries yield of several regional ecosystems among many things (Murawski 1993). Understanding the relationship between global warming and climate change as a result of anthropogenic activity is a vast and challenging process. In a bid to further this understanding, Rockström et al. (2009) introduced the concept of planetary boundaries. According to this concept, within each of these boundaries exists a level of safety that should not be exceeded. According to this concept, the planetary boundary for biodiversity loss, has been surpassed [@steffen]. Similarly, the safety boundary for warming of the globe by greenhouse gas emission has also been surpassed. The idea of planetary boundaries is essential in that it helps in understanding the threats humans cause via the continued destruction of the natural environment. Consequently, studies of anthropogenic influences on the ecosystem are generally divided into land or sea. Terrestrial ecosystems are more familiar to the general populace; whereas marine ecosystems are associated with high levels of primary production [@Chavez] and are highly at risk as a result of climate change [@osterblom]. Climate change has been proven to decrease primary production thus affecting trophic levels and leading to trophic turbulence and the destruction of marine fauna [@lewandowska; @dell, @mcCauley].

Climate change affects coastal species such as oysters, corals and, sea grass. Lord (2017) also found evidence that suggested that a change in climate as a result of rising sea temperatures may increase the spread of invasive species and increase the chances of an outbreak of diseases [@altizer]. Research by Mead et al (2013) found that coastal ecosystems are at high risk as a result of climate change, later, Whitfield et al (2016) confirmed that coastal bio-diversity has been observed to be affected. Given the importance of primary productivity, it was determined that the Benguela upwelling system within the South African region should become the focus of this thesis. Here the increase in global temperatures is one of the main focus points along with chlorophyll concentration. By making use of these two measurements, an analysis of how chlorophyll concentration varies with temperature changes could be conducted.

## Benguela
